High performance computing for global optimization problems

### 2018-05-17 2018/ITNT/ITNT2018.tex
Long version

\documentclass[a4paper]{jpconf}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document}
\title{High performance computing for global optimization problems}

@@

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
optimized function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate a nonuniform grid in the search domain, denser near the 
global minimizer and more sparse far away from this one. In the 
present paper, the results of scalability investigation of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system \cite{Globalizer} are presented.

@@

The parallel index algorithm, which can be applied to solving such problems 
with partially defined functions and which has been implemented in the 
Globalizer system, described in details in \cite{Strongin2013}. The main idea of the algorithm 
consists in the reduction of the initial multidimensional problem to a set of 
the nested subproblems of less dimensionality and the solving of these ones 
in parallel. The dimensionality reduction schemes utilized in the operation 
of the index algorithm are described briefly in the next section.

@@

The considered dimensionality reduction scheme juxtaposes a
multidimensional problem with lipschitzian functions to a univariate
problem where the corresponding functions satisfy the uniform H{\"o}lder
condition (see \cite{Strongin2000}), i.e.
\[
\left|g_i(y(x'))-g_i (y(x''))\right| \leq H_i \left|x'-x'' \right|^{1/N}, \; 
x',x''\in [0,1], \; 1\leq i \leq m+1.
\]
Here $N$ is the dimensionality of the initial multidimensional problem and
the coefficients $H_i$ are related with the Lipschitz constants $L_i$ of
the initial problem by the inequalities $H_i \leq 2L_i \sqrt{N+3}$.

@@

In the present study, this scheme has been applied for $M=2$, i.e. only one 
nesting level has been used
\[
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}{\left\{\varphi(y): \; 
g_i(y)\leq 0, \; 1 \leq i \leq m\right\}}.
\]

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Globalizer}. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{BarkalovGergelLebedev,BarkalovGergel} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

Let us conduct the investigation of the scalability of the parallel 
algorithm by solving a series of 100 test problems of the constrained global 
optimization. In work \cite{Gergel2017} the approach has been proposed, which allowed 
generating the constrained global optimization problems with the following 
properties:
\begin{itemize}
	\item one could control the size of the feasible domain $Q_{m+1}$ with 
respect to the whole domain $D$;
	\item the global minimizer of the objective function is known a priori 
taking into account the constraints;
	\item the global minimizer of the objective function without accounting 
for the constraints is out of the feasible domain $Q_{m+1}$  (with the 
purpose of simulating the behavior of the constraints and the objective 
function in the applied constrained optimization problems).
\end{itemize}

@@

%The results of experiments performed on a series of test problems of the constrained global optimization with different time of computing the the problem functions demonstrate the index global optimization algorithm combined with the block recursive scheme for the dimensionality reduction to provide a good speedup on the considered problem class. The directions of further investigations include the expansion of the number of algorithms used for solving the subproblems on the GPUs.

\section*{Acknowledgments}
This study was supported by the Russian Science Foundation, project No 16-11-10150.

@@

\bibitem{Globalizer}
Sysoyev A V, Barkalov K A, Sovrasov V V, Lebedev I G and Gergel V P 2017 \textit{Lecture Notes in Computer Science} \textbf{10421} 492--99

\bibitem{Jones}
Floudas C A and Pardalos P M 2009 \textit{The Encyclopedia of Optimization } (Heidelberg: Springer) p 725--35

\bibitem{Zilinskas}
Paulavi\v{c}ius R, \v{Z}ilinskas J and Grothey A 2011 \textit{Optim. Meth. Soft.} \textbf{26(3)} 487--98

\bibitem{Evtushenko2009}
Evtushenko Yu G, Malkova V U and Stanevichyus A A 2009 \textit{Comput. Math. Math. Phys.} \textbf{49(2)} 246--60.

\bibitem{Evtushenko2014}
Evtushenko Yu G and Posypkin M A 2014 \textit{Autom. Remote Control} \textbf{75(6)} 1025--40

\bibitem{Strongin2017}
Barkalov K A and Strongin R G 2017 \textit{Lecture Notes in Computer Science} \textit{10556} 18--33

@@

\bibitem{BarkalovLebedev}
Barkalov K A and Lebedev I G 2016 \textit{Communications in Computer and Information Science} \textbf{687} 224--35

@@

\bibitem{BarkalovGergelLebedev}
Barkalov K, Gergel V, Lebedev I 2015 \textit{Lecture Notes in Computer Science} \textbf{9251} 307--18

\bibitem{Gergel2017}
Gergel V 2017 \textit{Lecture Notes in Computer Science} \textbf{10556} 314--19

>>>>

\documentclass[a4paper]{jpconf}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

\begin{document}
\title{High performance computing for global optimization problems}

@@

\section{Introduction}
In the present paper, the constrained multiextremal optimization problems and 
the parallel methods for solving these ones are considered. The fact that the 
global extremum is an integral characteristic of a problem is an important 
feature of the multiextremal problems. Thus, its finding is related to the 
construction of a coverage of the search domain and the computing the 
optimized function values at all points of this coverage. The dimensionality 
affects the difficulty of solving the problems of considered class crucially: 
the computational costs grow with increasing this one exponentially, 
therefore, for solving such problems, the methods are required, which 
generate a nonuniform grid in the search domain, denser near the 
global minimizer and more sparse far away from this one. In the 
present paper, the results of scalability investigation of the parallel 
index global optimization algorithm developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Strongin2000,Strongin2013} and implemented in the Globalizer 
system \cite{Globalizer,Globalizer1} are presented.

@@

The main idea of the index algorithm which can be applied to solving such problems 
with partially defined functions consists in the reduction of the initial multidimensional problem to a set of 
the subproblems of less dimensionality and the solving of these ones 
in parallel. The index method and dimensionality reduction schemes are described briefly in the next sections.

\section{Parallel index method}

In this section we consider a one-dimensional multiextremal optimization problem
\[
\varphi(x^\ast)=\min{\left\{\varphi(x): x \in [a,b], \; g_i(x)\leq 0, \; 1 \leq i \leq m\right\}}.
\]
Let us describe a rules of parallel index method for the case when the algorithm carries out during one iteration $p>1$ trials simultaneously (each trial in a separate thread or process). Let $k(n)$ be the total number of trials, carried out after $n$ parallel iterations.

Suppose $n\geq 1$  iterations of the method have been carried out, in which trials were performed at $k=k(n)$ points $x^i,1\leq i \leq k$. Then points $x^{k+1},\dots,x^{k+p}$  of search trials of the next $n+1$ iteration are determined according to the following rules.

Rule 1. Renumber the points $x^1,...,x^k$ of the preceding trials by the
lower indices in ascending order of coordinate values, i.e.
\[
0=x_0<x_1<\dots <x_k<x_{k+1}=1,
\]
and juxtapose to them the values $z_i=g_\nu(x_i), \; \nu=\nu(x_i), \; 1
\leq i \leq k$, computed at these points. The points $x_0=0$ and
$x_{k+1}=1$ are introduced additionally, while the values $z_0$ and
$z_{k+1}$ are not defined.

Rule 2. Classify the indices $i, \; 1 \leq i \leq k$, of the trial points
according to the number of the problem constraints fulfilled at these
points, by constructing the sets
\[
I_\nu =\left\{i:1 \leq i \leq k, \; \nu=\nu(x_i) \right\}, \; 1 \leq \nu \leq m+1,
\]
containing the numbers of all the points $x_i, \; 1 \leq i \leq k$, with
the same values of $\nu$. The end points $x_0=0$ and $x_{k+1}=1$ are
interpreted as the ones having indices equal to zero. An additional set,
$I_0=\left\{0,k+1\right\}$, corresponds to them.

Determine the maximum value of the index:
\[
M=\max\left\{\nu(x_i), \; 1 \leq i \leq k \right \}.
\]

Rule 3. Compute the current lower estimates,
\begin{equation}\label{Rule_Mu}
\mu = \max\left\{ \frac{\left|z_i-z_j\right|}{ x_i - x_j }, \; i,j \in I_\nu, \; i>j \right\},
\end{equation}
for the unknown Lipschitz constants $L_\nu$ of the functions $g_\nu(x),1
\leq \nu \leq m+1$. If a set $I_\nu$ contains less than two elements, or
if $\mu_\nu$ is equal to zero, then assume $\mu_\nu=1$.

Rule 4. For all nonempty sets $I_\nu, \; 1 \leq \nu \leq m+1$, compute the
estimates
\[
z_\nu^\ast = \left\{
   \begin{array}{lr}
     -\epsilon_\nu, & \nu < M,\\
     \min\{ g_\nu(x_i): i\in I_\nu \}, & \nu = M,
   \end{array}
\right.
 \]
where the nonnegative numbers $(\epsilon_1,...,\epsilon_m)$ are parameters
of the algorithm.

Rule 5. For each interval ($x_{i-1},x_i), \; 1 \leq i \leq k+1,$ compute
the \textit{characteristics} $R(i)$ :
\[
R(i)=2\Delta_i-4\frac{z_i-z_\nu^\ast}{r_\nu \mu_\nu}, \; \nu=\nu(x_i)>\nu(x_{i-1}),
\]
\begin{equation}\label{Rule_R}
R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_\nu^2 \mu_\nu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z_\nu^\ast}{r_\nu \mu_\nu}, \;  \nu=\nu(x_i)=\nu(x_{i-1}),
\end{equation}
\[
R(i)=2\Delta_i-4\frac{z_{i-1}-z_\nu^\ast}{r_\nu \mu_\nu}, \; \nu=\nu(x_{i-1})>\nu(x_i),
\]
where $\Delta_i=x_i - x_{i-1}$. The values $r_\nu > 1, \; 1 \leq
\nu \leq m+1,$ are parameters of the algorithm. An appropriate selection
of $r_\nu$ allows to consider the product $r_\nu \mu_\nu$ as an estimate
of the Lipschitz constants $L_\nu, \; 1 \leq \nu \leq m+1$.

Rule 6. Arrange characteristics  $R(i), 1 \leq i \leq k+1$, in decreasing order 
\begin{equation}\label{Rule_Max}
R(t_1)\geq R(t_2)\geq \dots \geq R(t_{k}) \geq R(t_{k+1})
\end{equation}
and select $p$ maximum characteristics with interval numbers $t_j, 1\leq j \leq p$.

Rule 7. Carry out new trials at points $x^{k+j}\in(x_{t_j-1},x_{t_j}), 1\leq j\leq p$, calculated using the formulae
\[
x^{k+j} = \frac{x_{t_j} + x_{t_j-1}}{2}, \textrm{ if } \nu(x_{t{_j}-1}) \neq \nu(x_{t_j}),
\]
\begin{equation}\label{Rule_X}
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \frac{z_{t_j}-z_{t_j-1}}{2r_\nu \mu_\nu}, \textrm{ if } \nu=\nu(x_{t_j-1})=\nu(x_{t_j}).
\end{equation}

The algorithm terminates if the condition $\Delta_{t_j}<\epsilon$ is satisfied at least for one number $t_j, 1 \leq j \leq p$ ; $\epsilon>0$ is the predefined accuracy.

This method of parallelization has the following justification. The characteristics of intervals (\ref{Rule_R}) used in the index method can be considered as probability measures of the global minimum point location in these intervals. Inequalities (\ref{Rule_Max}) arrange intervals according to their characteristics, and trials are carried out in parallel in the first $p$ intervals with the largest probabilities.

@@

The considered dimensionality reduction scheme juxtaposes a
multidimensional problem with lipschitzian functions to a univariate
problem where the corresponding functions satisfy the uniform H{\"o}lder
condition (see \cite{Strongin2000}), i.e.
\begin{equation}\label{holder}
\left|g_i(y(x'))-g_i (y(x''))\right| \leq H_i \left|x'-x'' \right|^{1/N}, \; 
x',x''\in [0,1], \; 1\leq i \leq m+1.
\end{equation}
Here $N$ is the dimensionality of the initial multidimensional problem and
the coefficients $H_i$ are related with the Lipschitz constants $L_i$ of
the initial problem by the inequalities $H_i \leq 2L_i \sqrt{N+3}$.

Relation (\ref{holder}) allows modification of the parallel index algorithm described in previous section for solving multidimensional problems reduced to one-dimensional ones. For this purpose the intervals lengths $\Delta_i$ from the rules (\ref{Rule_Mu}) and (\ref{Rule_R}) of the index method are replaced with the lengths in a new metric
\[
\Delta_i = \left(x_i-x_{i-1}\right)^{1/N},
\]
and formula (\ref{Rule_X}) is replaced by the expression
\[
x^{k+j} = \frac{x_{t_j}+x_{t_j-1}}{2} - \frac{\mathrm{sign}(z_{t_j}-z_{t_j-1})}{2r_\nu}\left[\frac{\left|z_{t_j}-z_{t_j-1}\right|}{\mu_\nu}\right]^N, \; \nu=\nu(x_{t_j-1})=\nu(x_{t_j}).
\]

Various modifications of this algorithm and the corresponding theory of convergence are given in \cite{Strongin2000,Strongin2013,Gergel2005}.

@@

The number of vectors $M$ and the quantity of components in each vector $N_1, N_2,...,N_M$ are the parameters of the nested optimization scheme and can be used for the forming of the subproblems with necessary properties. For example, if  $M=N$ i.e. $u_i=y_i, 1\leq i \leq N,$ each nested subproblem is a one-dimensional one. If $M=1$, i.e. $u=u_1=y$ , the solving of the problem is equivalent to solving this one using a single evolvent mapping $[0,1]$ into $D$; the nested subproblems are absent.

In the present study, this scheme has been applied for $M=2$
\[
\min_{y \in D}{\left\{\varphi(y): \; g_i(y)\leq 0, \; 1 \leq i \leq 
m\right\}}= \min_{u_1\in D_1}\min_{u_2\in D_2}{\left\{\varphi(y): \; 
g_i(y)\leq 0, \; 1 \leq i \leq m\right\}},
\]
i.e. only one nesting level has been used.

@@

The recursive scheme of solving the global optimization problems has been 
implemented in the Globzlizer solver developed in Lobachevsky State 
University of Nizhny Novgorod \cite{Globalizer,Globalizer1}. The global search methods and various 
dimensionality reduction schemes make the algorithmic base for the 
Globalizer. The numerical experiments, the results of which are presented in 
\cite{BarkalovGergelLebedev,BarkalovGergel} demonstrate these methods, at least, are not worse than the well known 
methods applied for similar purposes, and even overcome these ones with 
respect to some parameters.

Let us conduct the investigation of the scalability of the parallel 
algorithm by solving a series of 100 test problems of the constrained global 
optimization. In work \cite{Gergel2017} the approach has been proposed, which allowed 
generating the constrained global optimization problems with the following 
properties:
\begin{itemize}
	\item one could control the size of the feasible domain $Q_{m+1}$ with 
respect to the whole search domain $D$;
	\item the global minimizer of the objective function is known a priori 
taking into account the constraints;
	\item the global minimizer of the objective function without accounting 
for the constraints is out of the feasible domain $Q_{m+1}$  (with the 
purpose of simulating the behavior of the constraints and the objective 
function in the applied constrained optimization problems).
\end{itemize}

@@

%The results of experiments performed on a series of test problems of the constrained global optimization with different time of computing the problem functions demonstrate the index global optimization algorithm combined with the block recursive scheme for the dimensionality reduction to provide a good speedup on the considered problem class. The directions of further investigations include the expansion of the number of algorithms used for solving the subproblems on the GPUs.

\section*{Acknowledgments}
This research was supported by the Russian Science Foundation, project No 16-11-10150 ``Novel efficient methods
and software tools for the time consuming decision making problems with using supercomputers of superior performance''.

\bigskip

@@

\bibitem{Globalizer}
Sysoyev A V, Barkalov K A, Sovrasov V V, Lebedev I G and Gergel V P 2017 \textit{Lect. Notes Comput. Science} \textbf{10421} 492--499

\bibitem{Globalizer1}
Gergel  V, Barkalov K and Sysoyev A 2018 \textit{Numer. Algebra Contr. Optim.} \textbf{8(1)} 47--62

\bibitem{Jones}
Floudas C A and Pardalos P M 2009 \textit{The Encyclopedia of Optimization } (Heidelberg: Springer) p 725--735

\bibitem{Zilinskas}
Paulavi\v{c}ius R, \v{Z}ilinskas J and Grothey A 2011 \textit{Optim. Meth. Soft.} \textbf{26(3)} 487--498

\bibitem{Evtushenko2009}
Evtushenko Yu G, Malkova V U and Stanevichyus A A 2009 \textit{Comput. Math. Math. Phys.} \textbf{49(2)} 246--260.

\bibitem{Evtushenko2014}
Evtushenko Yu G and Posypkin M A 2014 \textit{Autom. Remote Control} \textbf{75(6)} 1025--1040

\bibitem{Strongin2017}
Barkalov K A and Strongin R G 2017 \textit{Lect. Notes Comput. Science} \textit{10556} 18--33

@@

\bibitem{BarkalovLebedev}
Barkalov K A and Lebedev I G 2016 \textit{Commun. Comp. Inf. Science} \textbf{687} 224--235

\bibitem{Gergel2005}
Gergel V P and Strongin R G 2005 \textit{Lect. Notes Comput. Science} \textit{2763} 76--88

\bibitem{BarkalovGergelLebedev}
Barkalov K, Gergel V, Lebedev I 2015 \textit{Lect. Notes Comput. Science} \textbf{9251} 307--318

@@

\bibitem{Gergel2017}
Gergel V 2017 \textit{Lect. Notes Comput. Science} \textbf{10556} 314--319
